---
title: "MethylKit"
author: "Yaamini Venkataraman"
date: "09/12/2018"
output: github_document
---

In this file, I'll identify differentially methylated loci (DMLs) using `methylKit`.

# Prepare R Markdown file

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Install packages

```{r}
install.packages("devtools") #Install the devtools package
require(devtools) #Load devtools
```

```{r}
source("http://bioconductor.org/biocLite.R") #Source package from bioconductor
BiocManager::install("methylKit") #Install methylkit
```

```{r}
install_github("al2na/methylKit", build_vignettes = FALSE, repos = BiocInstaller::biocinstallRepos(), dependencies = TRUE) #Install more methylKit options
```

```{r}
require(methylKit) #Load methylkit
```

# Obtain session information

```{r}
sessionInfo()
```

# Download files from `gannet`

```{bash}
wget -r -l1 --no-parent -A_dedup.sorted.bam https://gannet.fish.washington.edu/spartina/2019-09-03-project-gigas-oa-meth/analyses/2019-09-03-Bismark/ #Download files from gannet. They will be stored in the same directory structure as they are online.
```

```{bash}
mv gannet.fish.washington.edu/spartina/2019-09-03-project-gigas-oa-meth/analyses/2019-09-03-Bismark/* . #Move all files from gannet folder to analyses folder
```

```{bash}
rm -r gannet.fish.washington.edu #Remove the empty gannet directory
```

# Process methylation data

```{r}
analysisFiles <- list("YRVA_dedup.sorted.bam",
                      "YRVL_dedup.sorted.bam") #Put all .bam files into a list for analysis.
```

```{r}
sample.IDs <- list("1", "2") #Create list of sample IDs
treatmentSpecification <- c(0, 1) #Specify which treatment the samples were from. 1 is the treatment (low pH) and 0 is the control (ambient pH)
```

I'll use `processBismarkAln` with `mincov = 2` to quickly process my files. I'll then use `filterByCoverage` to set minimum and maximum coverage thresholds.

```{r}
processedFiles <- processBismarkAln(location = analysisFiles, sample.id = sample.IDs, assembly = "v3", read.context = "CpG", mincov = 2, treatment = treatmentSpecification) #Process files for CpG methylation. Use 2x coverage for faster processing. Coverage will be adjusted later.
```

```{r}
processedFilteredFilesCov5 <- filterByCoverage(processedFiles, lo.count = 5, lo.perc = NULL, hi.count = 100, hi.perc = NULL) #filter processed files using lo.count and hi.count coverage thresholds. Coverage should be no less than 5 and should not exceed 100.
```

```{r}
processedFilteredFilesCov10 <- filterByCoverage(processedFiles, lo.count = 10, lo.perc = NULL, hi.count = 100, hi.perc = NULL) #filter processed files using lo.count and hi.count coverage thresholds. Coverage should be no less than 10 and should not exceed 100.
```

# Differentially methylated loci

## Obtain methylation and coverage plots

```{bash}
mkdir 2019-09-12-Loci-Analysis #Create directory to save plots
```


```{r}
nFiles <- length(sample.IDs) #Count number of samples
fileName <- data.frame("nameBase" = rep("2019-09-12-Loci-Analysis/2019-09-12-Percent-CpG-Methylation", times = nFiles),
                       "nameBase2" = rep("2019-09-12-Loci-Analysis/2019-09-12-Percent-CpG-Coverage", times = nFiles),
                       "sample.ID" = 1:2) #Create new dataframe for filenames
head(fileName) #Confirm dataframe creation
```

### 5x coverage

```{r}
fileName$actualFileName1 <- paste(fileName$nameBase, "-Filtered", "-5xCoverage", "-Sample", fileName$sample.ID, ".jpeg", sep = "") #Create a new column for the full filename for filtered + 5x coverage + specific sample's percent CpG methylation plot
fileName$actualFileName2 <- paste(fileName$nameBase2, "-Filtered", "-5xCoverage", "-Sample", fileName$sample.ID, ".jpeg", sep = "") #Create a new column for the full filename for filtered + 5x coverage + specific sample's percent CpG coverage plot
head(fileName) #Confirm column creation
```

```{r}
for(i in 1:nFiles) { #For each data file
  jpeg(filename = fileName$actualFileName1[i], height = 1000, width = 1000) #Save file with designated name
  getMethylationStats(processedFilteredFilesCov5[[i]], plot = TRUE, both.strands = FALSE) #Get %CpG methylation information
  dev.off() #Turn off plotting device
} #Plot and save %CpG methylation information
```

```{r}
for(i in 1:nFiles) { #For each data file
  jpeg(filename = fileName$actualFileName2[i], height = 1000, width = 1000) #Save file with designated name
  getCoverageStats(processedFilteredFilesCov5[[i]], plot = TRUE, both.strands = FALSE) #Get CpG coverage information
  dev.off() #Turn off plotting device
} #Plot and save CpG coverage information
```

### 10x coverage

```{r}
fileName$actualFileName3 <- paste(fileName$nameBase, "-Filtered", "-10xCoverage", "-Sample", fileName$sample.ID, ".jpeg", sep = "") #Create a new column for the full filename for filtered + 10x coverage + specific sample's percent CpG methylation plot
fileName$actualFileName4 <- paste(fileName$nameBase2, "-Filtered", "-10xCoverage", "-Sample", fileName$sample.ID, ".jpeg", sep = "") #Create a new column for the full filename for filtered + 10x coverage + specific sample's percent CpG coverage plot
head(fileName) #Confirm column creation
```

```{r}
for(i in 1:nFiles) { #For each data file
  jpeg(filename = fileName$actualFileName3[i], height = 1000, width = 1000) #Save file with designated name
  getMethylationStats(processedFilteredFilesCov5[[i]], plot = TRUE, both.strands = FALSE) #Get %CpG methylation information
  dev.off() #Turn off plotting device
} #Plot and save %CpG methylation information
```

```{r}
for(i in 1:nFiles) { #For each data file
  jpeg(filename = fileName$actualFileName4[i], height = 1000, width = 1000) #Save file with designated name
  getCoverageStats(processedFilteredFilesCov5[[i]], plot = TRUE, both.strands = FALSE) #Get CpG coverage information
  dev.off() #Turn off plotting device
} #Plot and save CpG coverage information
```

## Obtain clustering information

### 5x coverage

```{r}
methylationInformationFilteredCov5Destrand <- unite(processedFilteredFilesCov5, destrand = TRUE) #Combine all processed files into a single table. Use destrand = TRUE to separate strands
head(methylationInformationFilteredCov5Destrand) #Confirm unite
```

```{r}
clusteringInformationFilteredCov5Destrand <- clusterSamples(methylationInformationFilteredCov5Destrand, dist = "correlation", method = "ward", plot = FALSE) #Save cluster information as a new object
```

```{r}
jpeg(filename = "2019-09-12-Loci-Analysis/2019-09-12-Full-Sample-Pearson-Correlation-Plot-FilteredCov5Destrand.jpeg", height = 1000, width = 1000) #Save file with designated name
getCorrelation(methylationInformationFilteredCov5Destrand, plot = TRUE) #Understand correlation between methylation patterns in different samples
dev.off()
```

```{r}
jpeg(filename = "2019-09-12-Loci-Analysis/2019-09-12-Full-Sample-CpG-Methylation-Clustering-FilteredCov5Destrand.jpeg", height = 1000, width = 1000) #Save file with designated name
clusterSamples(methylationInformationFilteredCov5Destrand, dist = "correlation", method = "ward", plot = TRUE) #Cluster samples based on correlation coefficients
dev.off()
```

```{r}
jpeg(filename = "2019-09-12-Loci-Analysis/2019-09-12-Full-Sample-Methylation-PCA-FilteredCov5Destrand.jpeg", height = 1000, width = 1000) #Save file with designated name
PCASamples(methylationInformationFilteredCov5Destrand) #Run a PCA analysis on percent methylation for all samples
dev.off() #Turn off plotting device
```

```{r}
jpeg(filename = "2019-09-12-Loci-Analysis/2019-09-12-Full-Sample-Methylation-Screeplot-FilteredCov5Destrand.jpeg", height = 1000, width = 1000) #Save file with designated name
PCASamples(methylationInformationFilteredCov5Destrand, screeplot = TRUE) #Run the PCA analysis and plot variances against PC number in a screeplot
dev.off()
```

### 10x coverage

```{r}
methylationInformationFilteredCov10Destrand <- unite(processedFilteredFilesCov10, destrand = TRUE) #Combine all processed files into a single table. Use destrand = TRUE to separate strands
head(methylationInformationFilteredCov10Destrand) #Confirm unite
```

```{r}
clusteringInformationFilteredCov10Destrand <- clusterSamples(methylationInformationFilteredCov10Destrand, dist = "correlation", method = "ward", plot = FALSE) #Save cluster information as a new object
```

```{r}
jpeg(filename = "2019-09-12-Loci-Analysis/2019-09-12-Full-Sample-Pearson-Correlation-Plot-FilteredCov10Destrand.jpeg", height = 1000, width = 1000) #Save file with designated name
getCorrelation(methylationInformationFilteredCov10Destrand, plot = TRUE) #Understand correlation between methylation patterns in different samples
dev.off()
```

```{r}
jpeg(filename = "2019-09-12-Loci-Analysis/2019-09-12-Full-Sample-CpG-Methylation-Clustering-FilteredCov10Destrand.jpeg", height = 1000, width = 1000) #Save file with designated name
clusterSamples(methylationInformationFilteredCov10Destrand, dist = "correlation", method = "ward", plot = TRUE) #Cluster samples based on correlation coefficients
dev.off()
```

```{r}
jpeg(filename = "2019-09-12-Loci-Analysis/2019-09-12-Full-Sample-Methylation-PCA-FilteredCov10Destrand.jpeg", height = 1000, width = 1000) #Save file with designated name
PCASamples(methylationInformationFilteredCov10Destrand) #Run a PCA analysis on percent methylation for all samples
dev.off() #Turn off plotting device
```

```{r}
jpeg(filename = "2019-09-12-Loci-Analysis/2019-09-12-Full-Sample-Methylation-Screeplot-FilteredCov10Destrand.jpeg", height = 1000, width = 1000) #Save file with designated name
PCASamples(methylationInformationFilteredCov10Destrand, screeplot = TRUE) #Run the PCA analysis and plot variances against PC number in a screeplot
dev.off()
```

## Obtain differentially methylated loci

### 5x coverage

```{r}
differentialMethylationStatsFilteredCov5Destrand <- calculateDiffMeth(methylationInformationFilteredCov5Destrand) #Calculate differential methylation statistics based on treatment indication from processBismarkAln
head(differentialMethylationStatsFilteredCov5Destrand) #Look at differential methylation statistics
```

```{r}
write.csv(differentialMethylationStatsFilteredCov5Destrand, "2019-09-12-Loci-Analysis/2019-09-12-All-Loci-Methylation-Statistic-Filtered-Destrand-50-Cov5.csv") #Save DML background as a .csv
```

I will try a 50% difference (what I did with *C. virginica*), but I will also try a 75% and 100% difference. Since I only have two samples, I want to increase the stringency so I feel more confident in my DML.

```{r}
diffMethStats50FilteredCov5Destrand <- getMethylDiff(differentialMethylationStatsFilteredCov5Destrand, difference = 50, qvalue = 0.01) #Identify loci that are at least 50% different
head(diffMethStats50FilteredCov5Destrand) #Confirm creation
```

```{r}
write.csv(diffMethStats50FilteredCov5Destrand, "2019-09-12-Loci-Analysis/2019-09-12-Differentially-Methylated-Loci-Filtered-Destrand-50-Cov5.csv") #Save table as .csv
```

```{r}
diffMethStats75FilteredCov5Destrand <- getMethylDiff(differentialMethylationStatsFilteredCov5Destrand, difference = 75, qvalue = 0.01) #Identify loci that are at least 75% different
length(diffMethStats75FilteredCov5Destrand$chr) #Count the number of DML
head(diffMethStats75FilteredCov5Destrand) #Confirm creation
```

```{r}
diffMethStats100FilteredCov5Destrand <- getMethylDiff(differentialMethylationStatsFilteredCov5Destrand, difference = 99, qvalue = 0.001) #Identify loci that are at least 100% different
length(diffMethStats100FilteredCov5Destrand$chr) #Count the number of DML
head(diffMethStats100FilteredCov5Destrand) #Confirm creation
```

```{r}
write.csv(diffMethStats100FilteredCov5Destrand, "2019-09-12-Loci-Analysis/2019-09-12-Differentially-Methylated-Loci-Filtered-Destrand-100-Cov5.csv") #Save table as .csv
```

I'm going to use the following settings moving forward: 100% difference and q-value < 0.001. These are more stringent and narrow down the number of DML.

#### Hypermethylated DML

```{r}
diffMethStats50FilteredCov5DestrandHyper <- getMethylDiff(differentialMethylationStatsFilteredCov5Destrand, difference = 50, qvalue = 0.01, type = "hyper") #Identify hypermethylated loci that are at least 50% different
head(diffMethStats50FilteredCov5DestrandHyper) #Confirm creation
```

```{r}
write.csv(diffMethStats50FilteredCov5DestrandHyper, "2019-09-12-Loci-Analysis/2019-09-12-Differentially-Methylated-Loci-Filtered-Destrand-50-Cov5-Hypermethylated.csv") #Save table as .csv
```

```{r}
diffMethStats100FilteredCov5DestrandHyper <- getMethylDiff(differentialMethylationStatsFilteredCov5Destrand, difference = 99, qvalue = 0.001, type = "hyper") #Identify hypermethylated loci that are at least 50% different
length(diffMethStats100FilteredCov5DestrandHyper$chr) #Count number of DML
head(diffMethStats100FilteredCov5DestrandHyper) #Confirm creation
```

```{r}
write.csv(diffMethStats100FilteredCov5DestrandHyper, "2019-09-12-Loci-Analysis/2019-09-12-Differentially-Methylated-Loci-Filtered-Destrand-100-Cov5-Hypermethylated.csv") #Save table as .csv
```

#### Hypomethylated DML

```{r}
diffMethStats50FilteredCov5DestrandHypo <- getMethylDiff(differentialMethylationStatsFilteredCov5Destrand, difference = 50, qvalue = 0.01, type = "hypo") #Identify hypomethylated loci that are at least 50% different
head(diffMethStats50FilteredCov5DestrandHypo) #Confirm creation
```

```{r}
write.csv(diffMethStats50FilteredCov5DestrandHypo, "2019-09-12-Loci-Analysis/2019-09-12-Differentially-Methylated-Loci-Filtered-Destrand-50-Cov5-Hypomethylated.csv") #Save table as .csv
```

```{r}
diffMethStats100FilteredCov5DestrandHypo <- getMethylDiff(differentialMethylationStatsFilteredCov5Destrand, difference = 99, qvalue = 0.001, type = "hypo") #Identify hypomethylated loci that are at least 100% different
length(diffMethStats100FilteredCov5DestrandHypo$chr) #Count number of DML
head(diffMethStats100FilteredCov5DestrandHypo) #Confirm creation
```

```{r}
write.csv(diffMethStats100FilteredCov5DestrandHypo, "2019-09-12-Loci-Analysis/2019-09-12-Differentially-Methylated-Loci-Filtered-Destrand-100-Cov5-Hypomethylated.csv") #Save table as .csv
```

### 10x coverage

```{r}
differentialMethylationStatsFilteredCov10Destrand <- calculateDiffMeth(methylationInformationFilteredCov10Destrand) #Calculate differential methylation statistics based on treatment indication from processBismarkAln
head(differentialMethylationStatsFilteredCov10Destrand) #Look at differential methylation statistics
```

```{r}
write.csv(differentialMethylationStatsFilteredCov10Destrand, "2019-09-12-Loci-Analysis/2019-09-12-All-Loci-Methylation-Statistic-Filtered-Destrand-50-Cov10.csv") #Save DML background as a .csv
```

```{r}
diffMethStats50FilteredCov10Destrand <- getMethylDiff(differentialMethylationStatsFilteredCov10Destrand, difference = 50, qvalue = 0.01) #Identify loci that are at least 50% different
head(diffMethStats50FilteredCov10Destrand) #Confirm creation
```

```{r}
write.csv(diffMethStats50FilteredCov10Destrand, "2019-09-12-Loci-Analysis/2019-09-12-Differentially-Methylated-Loci-Filtered-Destrand-50-Cov10.csv") #Save table as .csv
```

```{r}
diffMethStats100FilteredCov10Destrand <- getMethylDiff(differentialMethylationStatsFilteredCov10Destrand, difference = 99, qvalue = 0.001) #Identify loci that are at least 50% different
length(diffMethStats100FilteredCov10Destrand$chr) #Count number of DML
head(diffMethStats100FilteredCov10Destrand) #Confirm creation
```

```{r}
write.csv(diffMethStats100FilteredCov10Destrand, "2019-09-12-Loci-Analysis/2019-09-12-Differentially-Methylated-Loci-Filtered-Destrand-100-Cov10.csv") #Save table as .csv
```

#### Hypermethylated DML

```{r}
diffMethStats50FilteredCov10DestrandHyper <- getMethylDiff(differentialMethylationStatsFilteredCov10Destrand, difference = 50, qvalue = 0.01, type = "hyper") #Identify hypermethylated loci that are at least 50% different
head(diffMethStats50FilteredCov10DestrandHyper) #Confirm creation
```

```{r}
write.csv(diffMethStats50FilteredCov10DestrandHyper, "2019-09-12-Loci-Analysis/2019-09-12-Differentially-Methylated-Loci-Filtered-Destrand-50-Cov10-Hypermethylated.csv") #Save table as .csv
```

```{r}
diffMethStats100FilteredCov10DestrandHyper <- getMethylDiff(differentialMethylationStatsFilteredCov10Destrand, difference = 99, qvalue = 0.001, type = "hyper") #Identify hypermethylated loci that are at least 50% different
length(diffMethStats100FilteredCov10DestrandHyper$chr) #Count number of DML
head(diffMethStats100FilteredCov10DestrandHyper) #Confirm creation
```

```{r}
write.csv(diffMethStats100FilteredCov10DestrandHyper, "2019-09-12-Loci-Analysis/2019-09-12-Differentially-Methylated-Loci-Filtered-Destrand-100-Cov10-Hypermethylated.csv") #Save table as .csv
```

#### Hypomethylated DML

```{r}
diffMethStats50FilteredCov10DestrandHypo <- getMethylDiff(differentialMethylationStatsFilteredCov10Destrand, difference = 50, qvalue = 0.01, type = "hypo") #Identify hypomethylated loci that are at least 50% different
head(diffMethStats50FilteredCov10DestrandHypo) #Confirm creation
```

```{r}
write.csv(diffMethStats50FilteredCov10DestrandHypo, "2019-09-12-Loci-Analysis/2019-09-12-Differentially-Methylated-Loci-Filtered-Destrand-50-Cov10-Hypomethylated.csv") #Save table as .csv
```

```{r}
diffMethStats100FilteredCov10DestrandHypo <- getMethylDiff(differentialMethylationStatsFilteredCov10Destrand, difference = 99, qvalue = 0.001, type = "hypo") #Identify hypomethylated loci that are at least 100% different
length(diffMethStats100FilteredCov10DestrandHypo$chr) #Count number of DML
head(diffMethStats100FilteredCov10DestrandHypo) #Confirm creation
```

```{r}
write.csv(diffMethStats100FilteredCov10DestrandHypo, "2019-09-12-Loci-Analysis/2019-09-12-Differentially-Methylated-Loci-Filtered-Destrand-100-Cov10-Hypomethylated.csv") #Save table as .csv
```

# Save DMLs as BEDfiles

## Install packages

**Do not** install these packages until analysis in `methylKit` is complete, as some of the installed packages will mask others important for analysis.

```{r}
require(readr) #Load package

install.packages("tidyverse") #Install tidyverse
require(tidyverse) #Load package
```

## DML Background

### 5x coverage

```{r}
head(methylationInformationFilteredCov5Destrand) #I only want the columns with chromosome, start position, stop position, and strand.
```

```{r}
methylationInformationFilteredCov5DestrandReduced <- data.frame("chr" = methylationInformationFilteredCov5Destrand$chr,
                                                                "start" = methylationInformationFilteredCov5Destrand$start,
                                                                "stop" = methylationInformationFilteredCov5Destrand$end,
                                                                "strand" = methylationInformationFilteredCov5Destrand$strand) #Subset file 
head(methylationInformationFilteredCov5DestrandReduced) #Confirm subsetting
```

```{r}
write_delim(methylationInformationFilteredCov5DestrandReduced, "2019-09-12-Methylation-Information-Filtered-Destrand-Cov5.bed", delim = "\t", col_names = FALSE) #Write out all methylation information as a background to be used for gene enrichment analyses.
```

```{bash}
head 2019-09-12-Methylation-Information-Filtered-Destrand-Cov5.bed #Confirm formatting was retained in export
```

### 10x coverage

```{r}
methylationInformationFilteredCov10DestrandReduced <- data.frame("chr" = methylationInformationFilteredCov10Destrand$chr,
                                                                "start" = methylationInformationFilteredCov10Destrand$start,
                                                                "stop" = methylationInformationFilteredCov10Destrand$end,
                                                                "strand" = methylationInformationFilteredCov10Destrand$strand) #Subset file 
head(methylationInformationFilteredCov10DestrandReduced) #Confirm subsetting
```

```{r}
methylationInformationFilteredCov10DestrandReduced$start <- (methylationInformationFilteredCov10DestrandReduced$start - 1) #Subtract 1 from the start position for visualization
head(methylationInformationFilteredCov10DestrandReduced) #Confirm subtracting worked
```

```{r}
write_delim(methylationInformationFilteredCov10DestrandReduced, "2019-09-12-Methylation-Information-Filtered-Destrand-Cov10.bed", delim = "\t", col_names = FALSE) #Write out all methylation information as a background to be used for gene enrichment analyses.
```

```{bash}
head 2019-09-12-Methylation-Information-Filtered-Destrand-Cov10.bed #Confirm formatting was retained in export
```

## DMLs

### 5x coverage and 50% difference

```{r}
DML5xFilteredDestrand09122019 <- mutate(diffMethStats50FilteredCov5Destrand, start = start - 3, end = end) %>% select(chr, start, end, meth.diff) %>% mutate_if(is.numeric, as.integer) #Save as a BED file, and avoid writing information in scientific notation
```

```{r}
head(DML5xFilteredDestrand09122019) #Confirm changes
```

```{r}
write_delim(DML5xFilteredDestrand09122019, "2019-09-12-DML-Destrand-5x-Locations.bed",  delim = '\t', col_names = FALSE) #Save data as a BED file
```

#### Hypermethylated DML

```{r}
DML5xFilteredDestrandHyper09122019 <- mutate(diffMethStats50FilteredCov5DestrandHyper, start = start -3, end = end) %>% select(chr, start, end, meth.diff) %>% mutate_if(is.numeric, as.integer) #Save as a BED file, and avoid writing information in scientific notation
```

```{r}
head(DML5xFilteredDestrandHyper09122019) #Confirm changes
```

```{r}
write_delim(DML5xFilteredDestrandHyper09122019, "2019-09-12-DML-Destrand-5x-Locations-Hypermethylated.bed",  delim = '\t', col_names = FALSE) #Save data as a BED file
```

#### Hypomethyalted DML

```{r}
DML5xFilteredDestrandHypo09122019 <- mutate(diffMethStats50FilteredCov5DestrandHypo, start = start - 3, end = end) %>% select(chr, start, end, meth.diff) %>% mutate_if(is.numeric, as.integer) #Save as a BED file, and avoid writing information in scientific notation
```

```{r}
head(DML5xFilteredDestrandHypo09122019) #Confirm changes
```

```{r}
write_delim(DML5xFilteredDestrandHypo09122019, "2019-09-12-DML-Destrand-5x-Locations-Hypomethylated.bed",  delim = '\t', col_names = FALSE) #Save data as a BED file
```

### 5x coverage and 100% difference

```{r}
DML5x100diffFilteredDestrand09122019 <- mutate(diffMethStats100FilteredCov5Destrand, start = start - 3, end = end) %>% select(chr, start, end, meth.diff) %>% mutate_if(is.numeric, as.integer) #Save as a BED file, and avoid writing information in scientific notation
head(DML5x100diffFilteredDestrand09122019) #Confirm changes
```

```{r}
write_delim(DML5x100diffFilteredDestrand09122019, "2019-09-12-DML-Destrand-5x-Locations-100diff.bed",  delim = '\t', col_names = FALSE) #Save data as a BED file
```

#### Hypermethylated DML

```{r}
DML5x100diffFilteredDestrandHyper09122019 <- mutate(diffMethStats100FilteredCov5DestrandHyper, start = start -3, end = end) %>% select(chr, start, end, meth.diff) %>% mutate_if(is.numeric, as.integer) #Save as a BED file, and avoid writing information in scientific notation
head(DML5x100diffFilteredDestrandHyper09122019) #Confirm changes
```

```{r}
write_delim(DML5x100diffFilteredDestrandHyper09122019, "2019-09-12-DML-Destrand-5x-Locations-Hypermethylated-100diff.bed",  delim = '\t', col_names = FALSE) #Save data as a BED file
```

#### Hypomethyalted DML

```{r}
DML5x100diffFilteredDestrandHypo09122019 <- mutate(diffMethStats100FilteredCov5DestrandHypo, start = start -3, end = end) %>% select(chr, start, end, meth.diff) %>% mutate_if(is.numeric, as.integer) #Save as a BED file, and avoid writing information in scientific notation
head(DML5x100diffFilteredDestrandHypo09122019) #Confirm changes
```

```{r}
write_delim(DML5x100diffFilteredDestrandHypo09122019, "2019-09-12-DML-Destrand-5x-Locations-Hypomethylated-100diff.bed",  delim = '\t', col_names = FALSE) #Save data as a BED file
```

### 10x coverage and 50% difference

```{r}
DML10xFilteredDestrand09122019 <- mutate(diffMethStats50FilteredCov10Destrand, start = start -3, end = end) %>% select(chr, start, end, meth.diff) %>% mutate_if(is.numeric, as.integer) #Save as a BED file, and avoid writing information in scientific notation
```

```{r}
head(DML10xFilteredDestrand09122019) #Confirm changes
```

```{r}
write_delim(DML10xFilteredDestrand09122019, "2019-09-12-DML-Destrand-10x-Locations.bed",  delim = '\t', col_names = FALSE) #Save data as a BED file
```

#### Hypermethylated DML

```{r}
DML10xFilteredDestrandHyper09122019 <- mutate(diffMethStats50FilteredCov10DestrandHyper, start = start -3, end = end) %>% select(chr, start, end, meth.diff) %>% mutate_if(is.numeric, as.integer) #Save as a BED file, and avoid writing information in scientific notation
```

```{r}
head(DML10xFilteredDestrandHyper09122019) #Confirm changes
```

```{r}
write_delim(DML10xFilteredDestrandHyper09122019, "2019-09-12-DML-Destrand-10x-Locations-Hypermethylated.bed",  delim = '\t', col_names = FALSE) #Save data as a BED file
```

#### Hypomethyalted DML

```{r}
DML10xFilteredDestrandHypo09122019 <- mutate(diffMethStats50FilteredCov10DestrandHypo, start = start -3, end = end) %>% select(chr, start, end, meth.diff) %>% mutate_if(is.numeric, as.integer) #Save as a BED file, and avoid writing information in scientific notation
```

```{r}
head(DML10xFilteredDestrandHypo09122019) #Confirm changes
```

```{r}
write_delim(DML10xFilteredDestrandHypo09122019, "2019-09-12-DML-Destrand-10x-Locations-Hypomethylated.bed",  delim = '\t', col_names = FALSE) #Save data as a BED file
```

### 10x coverage and 100% difference

```{r}
DML10x100diffFilteredDestrand09122019 <- mutate(diffMethStats100FilteredCov10Destrand, start = start -3, end = end) %>% select(chr, start, end, meth.diff) %>% mutate_if(is.numeric, as.integer) #Save as a BED file, and avoid writing information in scientific notation
head(DML10x100diffFilteredDestrand09122019) #Confirm changes
```

```{r}
write_delim(DML10x100diffFilteredDestrand09122019, "2019-09-12-DML-Destrand-10x-Locations-100diff.bed",  delim = '\t', col_names = FALSE) #Save data as a BED file
```

#### Hypermethylated DML

```{r}
DML10x100diffFilteredDestrandHyper09122019 <- mutate(diffMethStats100FilteredCov10DestrandHyper, start = start -3, end = end) %>% select(chr, start, end, meth.diff) %>% mutate_if(is.numeric, as.integer) #Save as a BED file, and avoid writing information in scientific notation
head(DML10x100diffFilteredDestrandHyper09122019) #Confirm changes
```

```{r}
write_delim(DML10x100diffFilteredDestrandHyper09122019, "2019-09-12-DML-Destrand-10x-Locations-Hypermethylated-100diff.bed",  delim = '\t', col_names = FALSE) #Save data as a BED file
```

#### Hypomethyalted DML

```{r}
DML10x100diffFilteredDestrandHypo09122019 <- mutate(diffMethStats100FilteredCov10DestrandHypo, start = start -1, end = end + 1) %>% select(chr, start, end, meth.diff) %>% mutate_if(is.numeric, as.integer) #Save as a BED file, and avoid writing information in scientific notation
head(DML10x100diffFilteredDestrandHypo09122019) #Confirm changes
```

```{r}
write_delim(DML10x100diffFilteredDestrandHypo09122019, "2019-09-12-DML-Destrand-10x-Locations-Hypomethylated-100diff.bed",  delim = '\t', col_names = FALSE) #Save data as a BED file
```
